\textit{MSF: Modulated Sub-graph Finder}

### 2018-07-31 MSF_Paper_FA.tex
changes and comments

\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{pdfpages}
\newcommand{\FA}[1]{\begingroup\color{magenta}#1\endgroup}
\newcommand{\TODO}[1]{\begingroup\color{red}#1\endgroup}
%opening
\begin{document}

@@

High throughput techniques such as RNA-seq or microarray analysis have
proven to be invaluable for the characterization of global transcriptional
gene activity changes due to external stimuli or diseases. Differential
gene expression analysis (DGEA) is the first step in the course of data
interpretation, typically producing lists of dozens to thousands of
differentially expressed genes. To further guide the interpretation of
these lists, different pathway analysis approaches have been
developed. These tools typically rely on the classification of genes into
set of genes, i.e. a pathway, based on the interactions between the genes
and their function in a common biological process. Regardless of technical
differences, these methods do not properly account for cross talk between
different pathways and rely on binary separation into differentially
expressed gene and unaffected genes based on an arbitrarily set p-value
cut-off.

@@

Existing pathway-based analysis approaches use different research designs,
which can be categorized into ORA (Over-representation analysis), FCS
(Functional class scoring) and pathway topology based methods. All of which
aim in finding in a subset of genes, e.g., significantly differentially
expressed genes, genes associated with a certain pathway more often than
expected given the total set of examined genes, e.g.~the whole genome
background.  ORA is the first and the most basic method of pathway
analysis. It uses a DEG list with user defined cut-off for the log-fold
change and \textit{p}-value (most commonly using absolute log-fold change
$\geq$ 2 and \textit{p}-value $\leq$ 0.05). Subsequently, sets of genes
associated with annotated pathways are tested for being over-represented in
the set of DEGs. To this end, hyper-geometric distribution, chi-square
tests, binomial probability or the Fisher’s exact test are used. Thereby
the information of the topology of genes in the pathways is
neglected~\cite{Bayer}. Furthermore, ORA assumes that the biological
pathways are independent of each other and ignores the fact that they
cross-talk and overlap~\cite{Khatri2012,Campos}.

Unlike ORA, FCS has no artificial cut-off to define DEG list. FCS works in
three step, first it calculates the gene-level statistics including
correlation of molecular measurements using differential expression of
individual genes, ANOVA, t-test and Z-score. In the second step the
statistics of individual genes in a pathway are transformed to an
individual pathway-level statistic commonly using Kolmogorov-Smirnov
statistic, mean or median. Finally the statistical significance of the
pathway-level statistics is assessed. Although FCS covers some of the
limitations from ORA, it still lacks the topology of genes in a pathway,
cross-talk and overlap of the pathways~\cite{Khatri2012,Campos}. Pathway
topology based methods are similar to FCS except that they consider the
topology of each gene during the gene-level statistics but still don't aim
to link different functional pathways~\cite{Khatri2012}.

@@

\section*{Methods}
\texttt{MSF} is developed as a novel heuristic approach to find concertedly
modulated sub-graphs in networks of biological interactions.  \texttt{MSF}
does not use predefined gene sets grouped into functional units, but rather
relies purely on the network of interacting genes.  The inputted network
consists of nodes corresponding to genes and edges representing
interactions. Furthermore it utilizes comprehensive results from a
differential gene expression analysis to discover the sub-graphs, or
modules, which are as a whole modulated.

@@

To reduce the complexity to score all possible connected sub-graphs
\texttt{MSF} applies a four steps heuristic as described in the
following. The proceeding identification of modulated sub-graphs from a
network by \texttt{MSF} is presented as a flowchart diagram
(Fig.~\ref{fig:pseudocodemsf}).

@@

\texttt{MSF} constructs the first sub-graph starting with the genes
associated with the lowest (most significant) \textit{p}-value deduced from
the DGEA. From this seed the sub-graph is attempted to be extended with its
neighboring genes, starting with the most significant one. A single
combined \textit{p}-value is calculated for the two genes. If the combined
\textit{p}-value is smaller than the minimal individual gene's
\textit{p}-value, the extended sub-graph is accepted. This step is
iteratively repeated until no further extension is accepted. In this case
the process starts over with all remaining genes not yet in a significantly
modulated sub-graph. This step identifies all the trivial sub-graphs that
are modulated in the whole network.\newline

@@

In the next step initial modulated sub-graphs are used to check if they
could further be extended beyond the immediate neighborhood. This is done
by testing all possible extension paths up to \emph{N} genes for all genes
in the sub-graph. Again, this step is iteratively repeated until no further
genes are added to the significant differentially expressed
sub-graphs. This step bridges small gaps of genes without a clear
differential signal in the DGEA.\newline

@@

After detection and extension of the modulated sub-graphs, they are tested
if combined sub-graphs score better than on their own. The merging of the
two sub-graphs is done by depth first search. If the two sub-graphs merge
with the connector of at most \emph{N} genes (default two genes) and the
combined \textit{p}-value of the merged sub-graph including the bridging
genes in between is less than the individual \textit{p}-values of the two
sub-graphs, the two sub-graphs are merged together to one big modulated
sub-graph. This step is repeated iteratively until no sub-graphs could be
merged.\newline

@@

To demonstrate its usefulness, \texttt{MSF} is applied to an RNA-seq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to the
Filoviridea family; filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host innate
and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now, there are no approved antiviral drugs
for the treatment of Ebola virus infection~\cite{Konde,Rhein}.  The initial
targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. It has been
proposed that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. Ebola infection
data was selected to test the approach because it has been well recognized for the last several decades, and vast literature is available for the pathogenesis of Ebola. Thereby facilitating the verification of the results of
\texttt{MSF} with the vast literature present on Ebola
infection. Especially, the detection of IFN-$\alpha / \beta$ as point of
action for the virus, could be considered as an basic indicator of the
rightness and usefulness of the approach.

@@

The EBOV infection experiments describe the course of infection at three
time-points 6, 24 and 48 hour post infection (hpi). For the earliest time
point at 6~hpi, five modulated sub-graphs were identified with 41, 107, and
18 genes. Two of the five sub-graphs were less than 4 genes long. Most of
the genes part of the sub-graphs were cytokines, chemokines (CXCL10, CCL8)
and Interleukin genes (IL6, IL27, IL23). IFNB1 and IFNA1 were both identified as two of the
possible sources. Most of the sources identified by \texttt{MSF} were
type~I interferon induced genes. At 24~hpi nine modulated sub-graphs were
identified with four main sub-graphs consisting
of 38, 85, 148 and 167 genes, others being smaller than 10 genes. IFNA1 was identified as one of the sources in
the most significantly modulated sub-graph. For the last time-point 48~hpi,
eleven modulated sub-graphs were identified. Eight of the sub-graphs were
less than seven genes and main sub-graphs had 39, 71 and 210 genes. IFNB1
and IFNA1 were identified as the two sources out of the total sources.

@@

 To assess the robustness and stability of our method, Poisson distributed
 noise was added to the read counts of the three time-points data set, used
 in the previous analysis. Then DGEA was carried out on the disturbed data
 with the same parameters as for the native data using edgeR, followed by
 analysis with \texttt{MSF}. This procedure was carried out 100 times.
 Every time the genes from the modulated sub-graphs identified from noisy
 data were compared to the genes of sub-graphs identified from the native
 data. This was also done for the DEG obtained for each run, using three different cutoffs of
 FDR 0.01, 0.05 and 0.1. The robustness of \texttt{MSF} and the DEG analysis
 for the time-point 6, 24, and 48~hpi are shown in
 Fig.~\ref{fig:DEGvsMSF}. The procedure how data noise was modeled can be
 considered as rather stringent, which is already reflected by the limited
 recall rate in the edgeR based DGEA, between 68~\% (6~hpi) and 93~\%
 (24~hpi). For \texttt{MSF}-analysis the observed median recall rates lay
 between 71~\% (6~hpi) and 84~\% (48~hpi). The better performance of the
 pure DGEA can be explained by the fact that disturbed p-values do not
 change the results for DGEA as long as the p-value does not wise above the
 chosen cutoff value. In contrast, \texttt{MSF} is sensitive to p-value
 changes across the whole range of possible values.

@@

The comparison shows most of the pathways known from literature
to be dis-regulated by Ebola infection are enriched in both the enrichment
analysis. Toll-Like receptor signaling pathway when interacts with EBOV
glycoprotein (GP), it triggers the activation of
cytokines~\cite{Olejnik}. Toll-like receptor pathway is expected to be dis-regulated in the early stage of infection,  this pathway was not identified as significantly dis-regulated when p-value
cut off DEG lists were analyzed for enrichment. Ten toll-like receptor cascades were identified as dis-regulated from gene enrichmant analysis of \texttt{MSF} identified genes, not a single one of these pathway was shown to be dis-regulated in DEG cut off lists. Since \texttt{MSF}
considers the complete DEG results, even the weak signal at the earliest
time-point was detected; for example Toll-like receptor signaling.

@@

Tlhe Ebola infection RNA-seq dataset analyzed during the current study are
available in the GEO repository (GSE84188)~\cite{Olejnik}. The cell
signaling network file used is from Reactome Functional interactions (FIs)
Version 2016~\cite{Cytokegg}

@@

This work was funded by the FWF (“Fonds zur Forderung der
wissenschaftlichen Forschung”) within the project Internationalen
Kooperationsprojektes - Intl cooperation Project (Joint Project - Lead
Agency Verfahren) with the project number (I 1988-B22). The grant was
assigned to ILH. FA was funded by the Austrian Science Fund (FWF) project
SFB F43.

@@

>>>>

\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{pdfpages}
\newcommand{\FA}[1]{\begingroup\color{magenta}#1\endgroup}
\newcommand{\ILH}[1]{\begingroup\color{blue}#1\endgroup}
\newcommand{\TODO}[1]{\begingroup\color{red}#1\endgroup}
%opening
\begin{document}

@@

High throughput techniques such as RNA-seq or microarray analysis have
proven to be invaluable for the characterization of global transcriptional
gene activity changes due to external stimuli or diseases. Differential
gene expression analysis (DGEA) is the first step in the course of data
interpretation, typically producing lists of dozens to thousands of
differentially expressed genes. To further guide the interpretation of
these lists, different pathway analysis approaches have been
developed. These tools typically rely on the classification of genes into
sets of genes, such as pathways, based on the interactions between the genes
and their function in a common biological process. Regardless of technical
differences, these methods do not properly account for cross talk between
different pathways and rely on binary separation into differentially
expressed gene and unaffected genes based on an arbitrarily set p-value
cut-off.

@@

Existing pathway-based analysis approaches use different research designs,
which can be categorized into ORA (Over-representation analysis), FCS
(Functional class scoring) and pathway topology based methods. All of which
aim to find a subset of genes, e.g., significantly differentially
expressed genes, genes associated with a certain pathway more often than
expected given the total set of examined genes, e.g.~the whole genome
background.  ORA \cite{} is the first and the most basic method of pathway
analysis. It uses a DEG list with user defined cut-off for the log-fold
change and \textit{p}-value (most commonly using absolute log-fold change
$\geq$ 2 and \textit{p}-value $\leq$ 0.05). Subsequently, sets of genes
associated with annotated pathways are tested for being over-represented in
the set of DEGs. To this end, hyper-geometric distribution, chi-square
tests, binomial probability or the Fisher’s exact test are used. Thereby
the information of the topology of genes in the pathways is
neglected~\cite{Bayer}. Furthermore, ORA assumes that the biological
pathways are independent of each other and ignores the fact that they
cross-talk and overlap~\cite{Khatri2012,Campos}.

Unlike ORA, FCS has no artificial cut-off to define a DEG list. FCS works in
three steps. First it calculates the gene-level statistics including
correlation of molecular measurements using differential expression of
individual genes, ANOVA, t-test and Z-score. In the second step the
statistics of individual genes in a pathway are transformed to an
individual pathway-level statistic commonly using Kolmogorov-Smirnov
statistic, mean or median. Finally the statistical significance of the
pathway-level statistics is assessed. Although FCS covers some of the
limitations from ORA, it still ignores the topology of genes in a pathway,
cross-talk and overlap of the pathways~\cite{Khatri2012,Campos}. Pathway
topology based methods are similar to FCS except that they consider the
topology of each gene during the gene-level statistics but still don't aim
to link different functional pathways~\cite{Khatri2012}.

@@

\section*{Methods}
\texttt{MSF} is developed as a novel heuristic approach to find concertedly
modulated sub-graphs in networks of biological interactions.  \texttt{MSF}
does not use predefined gene sets grouped into functional units, but rather
relies purely on the network of interacting genes.  The input network
consists of nodes corresponding to genes and edges representing
interactions. Furthermore it utilizes comprehensive results from a
differential gene expression analysis to discover the sub-graphs, or
modules, which are as a whole modulated.

@@

To reduce the complexity to score all possible connected sub-graphs
\texttt{MSF} applies a four step heuristic as described in the
following. The proceeding identification of modulated sub-graphs from a
network by \texttt{MSF} is presented as a flowchart diagram
(Fig.~\ref{fig:pseudocodemsf}).

@@

\texttt{MSF} constructs the first sub-graph starting with the genes
associated with the lowest (most significant) \textit{p}-value deduced from
the DGEA. From this seed it tries to extend the sub-graph by adding directly
neighboring genes, starting with the most significant one. A single
combined \textit{p}-value is calculated for the extended sub-graph.
If the combined \textit{p}-value is smaller than the \textit{p}-value of
the original sub-graph,
the extended sub-graph is accepted. This step is
iteratively repeated until no further extension is accepted. In this case
the process starts over with all remaining genes not yet in a significantly
modulated sub-graph. This step identifies all simple sub-graphs that
are modulated in the whole network.\newline

@@

In the next step, we check if any of the initial modulated sub-graphs 
could further be extended by adding more than one gene at a time. This is done
by testing all possible extension paths up to \emph{N} (usually 2) genes at
all nodes in the sub-graph. Again, this step is iteratively repeated until
no further genes are added to the significant differentially expressed
sub-graphs. This step bridges small gaps of genes without a clear
differential signal in the DGEA.\newline

@@

After detection and extension of the modulated sub-graphs, they are tested
if combined sub-graphs score better than on their own. The merging of the
two sub-graphs is done by depth first search \ILH{unclear}. If the two sub-graphs merge
with the connector of at most \emph{N} genes (default two genes) and the
combined \textit{p}-value of the merged sub-graph including the bridging
genes in between is less than the individual \textit{p}-values of the two
sub-graphs, the two sub-graphs are merged together to one big modulated
sub-graph. This step is repeated iteratively until no sub-graphs could be
merged.\newline

@@

To demonstrate its usefulness, \texttt{MSF} is applied to an RNA-seq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to the
Filoviridea family; filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host innate
and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now, there are no approved antiviral drugs
for the treatment of Ebola virus infection~\cite{Konde,Rhein}.  The initial
targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. It has been
proposed that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. Ebola infection
data was selected to test the approach because it has been well recognized for the last several decades, and vast literature is available for the pathogenesis of Ebola. Thereby facilitating the verification of the results of
\texttt{MSF} with the vast literature present on Ebola
infection. Especially, the detection of IFN-$\alpha / \beta$ as point of
action for the virus, could be considered as an basic indicator of the
correctness and usefulness of the approach.

@@

The EBOV infection experiments describe the course of infection at three
time-points 6, 24 and 48 hour post infection (hpi). For the earliest time
point at 6~hpi, three large modulated sub-graphs were identified with 41,
107, and 18 genes were identified, as well as two with less than 4 genes.
The modulated sub-graphs consist predominantly of cytokines, chemokines
(CXCL10, CCL8) and Interleukin genes (IL6, IL27, IL23).
IFNB1 and IFNA1 were both identified as two of the possible sources
\ILH{(in all subgraphs?)}.
Most of the sources identified by \texttt{MSF} were
type~I interferon induced genes \ILH{(link to some supplementary
  material with identified sub-graphs)}.
At 24~hpi nine modulated sub-graphs were
identified with four main sub-graphs consisting
of 38, 85, 148 and 167 genes, others being smaller than 10 genes. IFNA1 was identified as one of the sources in
the most significantly modulated sub-graph. For the last time-point 48~hpi,
eleven modulated sub-graphs were identified. Eight of the sub-graphs were
less than seven genes and main sub-graphs had 39, 71 and 210 genes. IFNB1
and IFNA1 were identified as \TODO{the two sources out of the total sources}.

@@

A potential concern is how noise in the gene expression measurements
affects our analysis. To assess the robustness and stability of our method,
we therefore added Poisson distributed noise
to the read counts of the three time-points data set, used above.
Then DGEA was carried out on the disturbed data
with the same parameters as for the native data using edgeR, followed by
analysis with \texttt{MSF}. This procedure was carried out 100 times.
Every time the genes from the modulated sub-graphs identified from noisy
data were compared to the genes of sub-graphs identified from the native
data. This was also done for the DEG obtained for each run, using three different cutoffs of
FDR 0.01, 0.05 and 0.1. The robustness of \texttt{MSF} and the DEG analysis
for the time-point 6, 24, and 48~hpi are shown in
Fig.~\ref{fig:DEGvsMSF}. The procedure how data noise was modeled can be
considered as rather stringent, which is already reflected by the limited
recall rate in the edgeR based DGEA, between 68~\% (6~hpi) and 93~\%
(24~hpi). For \texttt{MSF}-analysis the observed median recall rates lay
between 71~\% (6~hpi) and 84~\% (48~hpi). The better performance of the
pure DGEA can be explained by the fact that disturbed p-values do not
change the results for DGEA as long as the p-value does not rise above the
chosen cutoff value. In contrast, \texttt{MSF} is sensitive to p-value
changes across the whole range of possible values.

@@

The comparison shows most of the pathways known from literature
to be dis-regulated by Ebola infection are enriched in both the enrichment
analysis. Toll-Like receptor signaling pathway when interacts with EBOV
glycoprotein (GP), it triggers the activation of
cytokines~\cite{Olejnik}. Toll-like receptor pathway is expected to be dis-regulated in the early stage of infection,  this pathway was not identified as significantly dis-regulated when p-value
cut off DEG lists were analyzed for enrichment. Ten toll-like receptor cascades were identified as dis-regulated from gene enrichment analysis of \texttt{MSF} identified genes, not a single one of these pathway was shown to be dis-regulated in DEG cut off lists. Since \texttt{MSF}
considers the complete DEG results, even the weak signal at the earliest
time-point was detected; for example Toll-like receptor signaling.

@@

The Ebola infection RNA-seq data set analyzed during the current study are
available in the GEO repository (GSE84188)~\cite{Olejnik}. The cell
signaling network file used is from Reactome Functional interactions (FIs)
Version 2016~\cite{Cytokegg}

@@

This work was funded by the FWF (“Fonds zur F{\"o}rderung der
wissenschaftlichen Forschung”) within the project Internationalen
Kooperationsprojektes - Intl cooperation Project (Joint Project - Lead
Agency Verfahren) with the project number (I 1988-B22). The grant was
assigned to ILH. FA was funded by the Austrian Science Fund (FWF) project
SFB F43.

@@

% LocalWords:  MSF Mariam Farman Ivo Perutz microarray DGEA differentially
% LocalWords:  effectors DEGs KEGG Reactome STAT1 FCS ANOVA Kolmogorov IFN
% LocalWords:  Smirnov regulon StringApp Cytoscape transduction github FIs
% LocalWords:  vicinal Hartung Hartung's iteratively monocyte macrophages
% LocalWords:  MDMs GSE84188 EBOV Filoviridea dendritic pathogenesis edgeR
% LocalWords:  hpi cytokines chemokines CXCL10 CCL8 IL6 IL27 IL23 IFNB1
% LocalWords:  IFNA1 subgraphs INFB1 cytokine chemokine jak TNF apoptosis
% LocalWords:  ECM glycoprotein cytosolic INFA1 apoptotic Reactome's ILH
% LocalWords:  DESeq2 MRF FWF Fonds zur Forderung der wissenschaftlichen
